---
title: "F1 Analysis (2001–2024) — R/Tidyverse"
author: "Your Name"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    df_print: paged
---

> **Note**: This RMD expects the Kaggle *Formula 1 World Championship* CSVs (Rao dataset) in the working directory:
> `races.csv`, `results.csv`, `drivers.csv`, `constructors.csv`, `qualifying.csv`, `pit_stops.csv`, `status.csv`, `constructor_standings.csv`, `driver_standings.csv`, `circuits.csv`.
> If you have a newer dataset (e.g., 1950–2024), the code will automatically include newer seasons.

```{r setup, message=FALSE, warning=FALSE}
library(readr)
library(dplyr)
library(tidyr)
library(stringr)
library(lubridate)
library(ggplot2)
library(forcats)
library(scales)
library(knitr)
```

## 1) Load data

```{r load-data}
# Change path if your CSVs are stored elsewhere
races            <- read_csv("races.csv", show_col_types = FALSE)
results          <- read_csv("results.csv", show_col_types = FALSE)
drivers          <- read_csv("drivers.csv", show_col_types = FALSE)
constructors     <- read_csv("constructors.csv", show_col_types = FALSE)
qualifying       <- suppressWarnings(read_csv("qualifying.csv", show_col_types = FALSE))
pit_stops        <- suppressWarnings(read_csv("pit_stops.csv", show_col_types = FALSE))
status           <- read_csv("status.csv", show_col_types = FALSE)
constructor_stds <- read_csv("constructor_standings.csv", show_col_types = FALSE)
driver_stds      <- read_csv("driver_standings.csv", show_col_types = FALSE)
circuits         <- read_csv("circuits.csv", show_col_types = FALSE)

# Harmonise dates
races <- races %>% mutate(raceDate = ymd(date))
drivers <- drivers %>% mutate(dob = ymd(dob))

# Identify available season range
season_min <- min(races$year, na.rm = TRUE)
season_max <- max(races$year, na.rm = TRUE)
season_min; season_max
```

### Filter analysis window

```{r filter-window}
# Focus window inspired by the referenced Kaggle notebook
analysis_start <- 2001
analysis_end   <- min(2024, season_max)  # handle whatever the dataset provides
races_win      <- races %>% filter(year >= analysis_start, year <= analysis_end)

# Handy joins
results_full <- results %>%
  inner_join(races %>% select(raceId, year, round, raceDate, circuitId, name), by = "raceId") %>%
  left_join(drivers %>% mutate(driverName = str_squish(paste(forename, surname))), by = "driverId") %>%
  left_join(constructors %>% rename(constructorName = name), by = "constructorId") %>%
  left_join(status, by = "statusId") %>%
  left_join(circuits %>% select(circuitId, circuitName = name, country), by = "circuitId")

# Helper: winner & podium flags
results_full <- results_full %>%
  mutate(
    is_win   = positionOrder == 1,
    is_podium= positionOrder %in% 1:3
  ) %>%
  filter(year >= analysis_start, year <= analysis_end)
```

## 2) Season-level views

### 2.1 Top drivers by wins (2001–present)

```{r top-drivers-wins, fig.width=8, fig.height=5}
top_drivers <- results_full %>%
  filter(is_win) %>%
  count(driverName, sort = TRUE) %>%
  slice_head(n = 15) %>%
  mutate(driverName = fct_reorder(driverName, n))

ggplot(top_drivers, aes(x = driverName, y = n)) +
  geom_col() +
  coord_flip() +
  labs(title = "Top 15 Drivers by Wins", x = NULL, y = "Wins since 2001")
```

### 2.2 Constructor dominance by season (wins per season)

```{r constructor-dominance, fig.width=9, fig.height=5}
constructor_wins_by_season <- results_full %>%
  filter(is_win) %>%
  count(year, constructorName)

ggplot(constructor_wins_by_season, aes(x = year, y = n, fill = constructorName)) +
  geom_col(position = "stack") +
  labs(title = "Constructor Wins per Season", x = "Season", y = "Wins") +
  theme(legend.position = "bottom")
```

### 2.3 Podium share by constructor (stacked %)

```{r podium-share, fig.width=9, fig.height=5}
podium_share <- results_full %>%
  filter(is_podium) %>%
  count(year, constructorName) %>%
  group_by(year) %>%
  mutate(pct = n / sum(n))

ggplot(podium_share, aes(x = year, y = pct, fill = constructorName)) +
  geom_col(position = "fill") +
  scale_y_continuous(labels = percent) +
  labs(title = "Podium Share by Constructor", x = "Season", y = "Share of Podiums") +
  theme(legend.position = "bottom")
```

## 3) Qualifying vs Race: how much does grid matter?

```{r grid-finish-heat, fig.width=8, fig.height=6}
grid_finish <- results_full %>%
  filter(!is.na(grid), grid >= 0, !is.na(positionOrder)) %>%
  mutate(grid_cut = case_when(
    grid <= 3  ~ "P1–P3",
    grid <= 6  ~ "P4–P6",
    grid <= 10 ~ "P7–P10",
    grid <= 15 ~ "P11–P15",
    TRUE       ~ "P16+"
  ),
  finish_cut = case_when(
    positionOrder <= 3  ~ "P1–P3",
    positionOrder <= 6  ~ "P4–P6",
    positionOrder <= 10 ~ "P7–P10",
    positionOrder <= 15 ~ "P11–P15",
    TRUE                ~ "P16+"
  )) %>%
  count(grid_cut, finish_cut) %>%
  group_by(grid_cut) %>%
  mutate(pct = n/sum(n))

ggplot(grid_finish, aes(x = grid_cut, y = finish_cut, fill = pct)) +
  geom_tile() +
  scale_fill_continuous(labels = percent) +
  labs(title = "Grid Position vs Finish — Heatmap", x = "Grid (buckets)", y = "Finish (buckets)", fill = "Share")
```

### Position change distribution

```{r position-change, fig.width=8, fig.height=5}
pos_change <- results_full %>%
  filter(grid >= 0, !is.na(positionOrder)) %>%
  mutate(delta = grid - positionOrder)

ggplot(pos_change, aes(x = delta)) +
  geom_histogram(binwidth = 1, boundary = 0) +
  labs(title = "Race Position Change (Grid - Finish)", x = "Positions Gained (+) / Lost (−)", y = "Count")
```

## 4) Reliability: DNF trends over time

```{r dnf-trends, fig.width=9, fig.height=5}
dnf_status_terms <- c("Accident","Collision","Engine","Gearbox","Hydraulics","Electrical","Brakes","Suspension","Transmission",
                      "Puncture","Overheating","Fuel","Mechanical","Clutch","Turbo","Driveshaft","Oil","Power Unit","DNF","Did not finish","Withdrew","Water pump")
dnf_by_year <- results_full %>%
  mutate(is_dnf = case_when(
    str_detect(tolower(status), "finished") ~ FALSE,
    str_detect(tolower(status), "lap") ~ FALSE, # finished but lapped
    TRUE ~ TRUE
  )) %>%
  group_by(year) %>%
  summarise(dnf_rate = mean(is_dnf, na.rm = TRUE))

ggplot(dnf_by_year, aes(x = year, y = dnf_rate)) +
  geom_line() + geom_point() +
  scale_y_continuous(labels = percent) +
  labs(title = "DNF Rate by Season", x = "Season", y = "DNF Rate")
```

## 5) Pit stops: how fast are they getting?

```{r pit-stops, fig.width=9, fig.height=5}
if (exists("pit_stops") && nrow(pit_stops) > 0) {
  pit_join <- pit_stops %>%
    left_join(races %>% select(raceId, year, raceDate), by = "raceId") %>%
    filter(year >= analysis_start, year <= analysis_end) %>%
    mutate(duration_ms = ifelse(!is.na(milliseconds), milliseconds, as.numeric(ms(duration))),
           duration_s = duration_ms / 1000)

  pit_summary <- pit_join %>%
    group_by(year) %>%
    summarise(median_pit = median(duration_s, na.rm = TRUE),
              p10 = quantile(duration_s, 0.1, na.rm = TRUE),
              p90 = quantile(duration_s, 0.9, na.rm = TRUE))

  ggplot(pit_summary, aes(x = year, y = median_pit)) +
    geom_line() + geom_point() +
    geom_ribbon(aes(ymin = p10, ymax = p90), alpha = 0.2) +
    labs(title = "Pit Stop Duration (median with 10–90% band)", x = "Season", y = "Seconds")
} else {
  plot.new(); text(0.5,0.5,"pit_stops.csv not available in this dataset")
}
```

## 6) Driver age & experience

```{r age-experience, fig.width=9, fig.height=5}
driver_debuts <- results_full %>%
  group_by(driverId) %>%
  summarise(first_year = min(year, na.rm = TRUE)) %>%
  left_join(drivers %>% select(driverId, driverName, dob), by = "driverId")

winners <- results_full %>% filter(is_win) %>%
  select(driverId, driverName, raceDate) %>%
  left_join(driver_debuts, by = "driverId") %>%
  mutate(age_win = as.numeric(difftime(raceDate, dob, units = "days"))/365.25,
         exp_years = year(raceDate) - first_year)

ggplot(winners, aes(x = year(raceDate), y = age_win)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "loess", se = FALSE) +
  labs(title = "Age of Race Winners Over Time", x = "Season", y = "Age at Win (years)")
```

## 7) Circuits: which tracks see the most overtaking (proxy)?

*Proxy:* positions gained (grid - finish) as a crude overtaking indicator.

```{r circuits-overtake, fig.width=9, fig.height=6}
circuit_overtake <- results_full %>%
  filter(grid >= 0, !is.na(positionOrder)) %>%
  group_by(circuitName, country) %>%
  summarise(mean_gain = mean(grid - positionOrder, na.rm = TRUE),
            races = n()) %>%
  filter(races >= 100) %>%
  arrange(desc(mean_gain)) %>%
  slice_head(n = 15) %>%
  mutate(label = fct_reorder(paste0(circuitName, " (", country, ")"), mean_gain))

ggplot(circuit_overtake, aes(x = label, y = mean_gain)) +
  geom_col() +
  coord_flip() +
  labs(title = "Circuits with Highest Avg. Positions Gained (proxy)", x = NULL, y = "Avg. Positions Gained")
```

## 8) Simple predictive model: podium probability

```{r model-podium, message=FALSE, warning=FALSE}
library(rsample)
library(recipes)
library(parsnip)
library(workflows)
library(yardstick)

model_df <- results_full %>%
  filter(!is.na(grid), grid >= 0, !is.na(positionOrder)) %>%
  mutate(podium = factor(if_else(positionOrder <= 3, "yes", "no")),
         grid_log = log1p(grid)) %>%
  group_by(driverId) %>%
  mutate(driver_form = zoo::rollapply(positionOrder, width = 5, align = "right", fill = NA, FUN = function(x) mean(x, na.rm = TRUE))) %>%
  ungroup() %>%
  mutate(driver_form = replace_na(driver_form, mean(driver_form, na.rm = TRUE))) %>%
  select(podium, grid, grid_log, year, constructorName, driver_form)

set.seed(42)
split <- initial_split(model_df, prop = 0.8, strata = podium)
train <- training(split)
test  <- testing(split)

rec <- recipe(podium ~ grid_log + year + constructorName + driver_form, data = train) %>%
  step_string2factor(constructorName) %>%
  step_dummy(all_nominal_predictors()) %>%
  step_zv(all_predictors())

mod <- logistic_reg() %>%
  set_engine("glm")

wf <- workflow() %>% add_model(mod) %>% add_recipe(rec)
fit <- fit(wf, data = train)

pred <- predict(fit, new_data = test, type = "prob") %>% bind_cols(test)
roc_auc(truth = podium, .pred_yes = .pred_yes)
```

```{r model-plot, fig.width=8, fig.height=5}
pred_cls <- predict(fit, new_data = test, type = "class") %>% bind_cols(pred)
cm <- conf_mat(pred_cls, truth = podium, estimate = .pred_class)
autoplot(cm, type = "heatmap") + labs(title = "Confusion Matrix — Podium Classifier")
```

## 9) Save useful tables

```{r save-csv, eval=FALSE}
write_csv(top_drivers, "top_drivers_since2001.csv")
write_csv(constructor_wins_by_season, "constructor_wins_by_season.csv")
write_csv(podium_share, "podium_share_by_season.csv")
```

## 10) Repro notes

- Data source: Kaggle — Formula 1 World Championship dataset (Rao), or newer forks.
- If you include seasons up to 2024, keep the same CSV schema and the code will adapt automatically.
- This RMD is designed to knit without internet access as long as CSVs are present.

```{r save-csv, eval=FALSE}
# Load necessary datasets
drivers <- read.csv('drivers.csv')
driver_standings <- read.csv('driver_standings.csv')
races <- read.csv('races.csv')
circuits <- read.csv('circuits.csv')

# Get Verstappen's driverId
verstappen_id <- drivers$driverId[tolower(drivers$surname) == 'verstappen']

# Filter driver's results and merge race and circuit info
verstappen_results <- subset(driver_standings, driverId == verstappen_id)
verstappen_race_results <- merge(verstappen_results, races[, c('raceId', 'circuitId', 'year')], by = 'raceId')
verstappen_race_results <- merge(verstappen_race_results, circuits[, c('circuitId', 'name')], by = 'circuitId')

# Calculate average finishing position by circuit
library(dplyr)
circuit_performance <- verstappen_race_results %>%
  group_by(name) %>%
  summarise(avg_position = mean(position, na.rm = TRUE)) %>%
  arrange(desc(avg_position))

head(circuit_performance, 10)

```





Performance Consistency by Season and Circuit
```{r}
library(dplyr)
library(ggplot2)

# Load files
drivers <- read.csv('drivers.csv')
driver_standings <- read.csv('driver_standings.csv')
races <- read.csv('races.csv')

# Get Max Verstappen's driverId
verstappen_id <- drivers$driverId[tolower(drivers$forename) == 'max' & tolower(drivers$surname) == 'verstappen']

# Join standings with races on raceId for year
verstappen_year_results <- driver_standings %>%
  filter(driverId == verstappen_id) %>%
  left_join(races %>% select(raceId, year), by='raceId')

# Calculate mean and sd for each year
yearly_consistency <- verstappen_year_results %>%
  group_by(year) %>%
  summarise(mean_position = mean(position, na.rm=TRUE),
            sd_position = sd(position, na.rm=TRUE),
            races = n())

print(yearly_consistency)

# Plot yearly consistency
ggplot(yearly_consistency, aes(x=year)) +
  geom_line(aes(y=mean_position), color='blue', size=1.2) +
  geom_point(aes(y=mean_position), color='blue', size=2) +
  geom_errorbar(aes(ymin = mean_position - sd_position,
                    ymax = mean_position + sd_position), width=0.2, color='gray') +
  labs(y='Average Finish Position', x='Year',
       title='Max Verstappen Yearwise Performance Consistency') +
  theme_minimal()

```

```{r save-csv, eval=FALSE}
library(dplyr)
library(ggplot2)

# Assume verstappen_year_results is already created as before
# Calculate mean and sd for each year
yearly_consistency <- verstappen_year_results %>%
  group_by(year) %>%
  summarise(mean_position = mean(position, na.rm=TRUE),
            sd_position = sd(position, na.rm=TRUE),
            races = n())

print(yearly_consistency) # Table output

# Improved Plot
ggplot(yearly_consistency, aes(x=year, y=mean_position)) +
  geom_line(color='blue', size=1.5) +
  geom_point(color='blue', size=3) +
  geom_errorbar(aes(ymin=mean_position - sd_position, ymax=mean_position + sd_position), width=0.2, color='gray') +
  geom_text(aes(label=round(mean_position,1)), vjust=-1, size=5, color='darkblue') +
  scale_y_reverse(breaks=seq(1, max(yearly_consistency$mean_position), by=1)) + # Lower positions are better
  labs(title='Max Verstappen Annual Performance Consistency',
       subtitle='Average race finish position per year (lower is better)',
       x='Year',
       y='Average Finish Position') +
  theme_minimal(base_size = 15) +
  theme(panel.grid.major = element_line(color='gray80'),
        plot.title = element_text(face='bold', hjust=0.5),
        plot.subtitle = element_text(hjust=0.5)) 


```
The analysis began by extracting Max Verstappen's race performance from 2015 to 2024 using the F1 historical datasets. First, his race results were joined with season and race information to group his finishing positions by year. Summary statistics for each season were then calculated, specifically the average (mean_position) and standard deviation (sd_position) of his finishing places, as well as the number of races per year. These metrics were visualized in an improved line plot, where lower average positions correspond to better race results and smaller error bars indicate higher consistency. The results clearly show a dramatic improvement in Verstappen's average finish, dropping from nearly 12th place in 2015 to a consistent 1st place from 2023 onwards, reflecting both increasing performance and consistency across seasons. This pattern demonstrates Verstappen's growth from a midfield driver to a dominant, championship-winning performer, substantiated by both the summary table and the visual trend in the graph.

```{r save-csv, eval=FALSE}


```

```{r save-csv, eval=FALSE}


```

```{r save-csv, eval=FALSE}


```

```{r save-csv, eval=FALSE}


```

```{r save-csv, eval=FALSE}


```

```{r save-csv, eval=FALSE}


```

```{r save-csv, eval=FALSE}


```

```{r save-csv, eval=FALSE}


```

```{r save-csv, eval=FALSE}


```

```{r save-csv, eval=FALSE}


```

```{r save-csv, eval=FALSE}


```